### 2018-01-15 paper/dispRity_paper_highlighted.tex
paper up

The time-binning method groups elements by specific age range.
The time-slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
When a slice occurs not on a tip or a node, \hl{six} methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
\begin{itemize}
    \item ``acctran'' where the data chosen along the branch is always the one of the descendant
    \item ``deltran'' where the data chosen along the branch is always the one of the ancestor
    \item \hl{``random''} where the data chosen along the branch is randomly chosen between the descendant or the ancestor
    \item \hl{``proximity''} where the data chosen along the branch is either the descendant or the ancestor depending on branch length
    \item \hl{``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
    \item \hl{``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
\end{itemize}

\hl{Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).
See Guillerme \& Cooper \textit{in review} for full description.}

\subsection{Bootstrapping and rarefying}
Disparity measurement can be influenced by sampling \citep{Butler2012}.
To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different subsets).
Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data to obtain a distribution on which to perform statistical analyses.

Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).

Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of subsets with the same number of elements in each.
In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).

\subsection{Interpreting results}
The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
These results can also be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.

\subsubsection{Summarising and plotting}
The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity-through-time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e. non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for a certain number of elements (i.e. the rarefaction level).
The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.

\subsubsection{Testing hypotheses}
The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data (e.g. by comparing the different subsets to each other).
Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsets: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subset to all the others; (3) ``sequential'' for comparing subsets sequentially (e.g. first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e. \texttt{disparity $\mathtt{\sim}$ subsets}) or (5) any list of pairs (or more) of subsets to compare.

@@

\section{Examples}
Multivariate analysis can be really useful for looking
 \hl{at} 
 multiple aspects of organisms diversity together.
For example, one classic univariate trait used in ecology and evolution can be the number of species present in a ecosystem or/and through time.
This trait can be useful to test whether some ecosystems or/and time periods have more species diversity than others.
In this case one can then test some hypothesis such as: environments with
 \hl{longer altitudinal} 
 gradients contain more species.

Conversely, in the case of species diversity, one can also look at another independent multivariate trait: the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
Following the previous example, if disparity is also higher with
 \hl{longer altitudinal} 
 gradients, one could argue that the species richness is caused by an increased number of morphologies adapted to more ecological niches (or something akin).

The following example is based on a classical morphological disparity analysis.
Note that more examples are available in the package manual (\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).

@@

In this example, the space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace (i.e. a large sum of variances approximates a large morphospace and vice versa).

\subsection{Example}

\hl{The following example is based on the discrete morphological data described above. More examples, including based on other types of data are available in the package vignette.}

>>>>

The time-binning method groups elements by specific age range.
The time-slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
When a slice occurs not on a tip or a node, \hl{six} methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice.
\hl{For each method, the selected data is:}
\begin{itemize}
    \item ``acctran'': always the descendant
    \item ``deltran'': always the ancestor
    \item \hl{``random''}: randomly chosen between the descendant or the ancestor
    \item \hl{``proximity''}: either the descendant or the ancestor depending on branch length
    \item \hl{``equal.split'': both the descendant and the ancestor with an even probability}
    \item \hl{``gradual.split'': both the descendant and the ancestor with a probability depending on branch length}
\end{itemize}

\hl{Note that the four first models are a proxy for punctuated evolution and the two last ones for gradual evolution.
There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.
See Guillerme \& Cooper \textit{in review} for full description.}

\subsection{Bootstrapping and rarefying}
Disparity measurement can be influenced by sampling \citep{Butler2012}.
To take this source of bias into account, one can bootstrap the multidimensional space or/and to rarefy the data.
Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data to obtain a distribution on which to perform statistical analyses.

Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' algorithm where \hl{$n = 1$} (somewhat akin to jackknife resampling

Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).

\subsection{Interpreting results}
The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity}.
These results can also be analysed using the \texttt{test.dispRity} function for comparing subsets or testing hypotheses.

\subsubsection{Summarising and plotting}
The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e. non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for a certain number of elements (i.e. the rarefaction level).
The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.

\subsubsection{Testing hypotheses}
The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data.
Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsets: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subset to all the others; (3) ``sequential'' for comparing subsets sequentially (e.g. first against second, second against third, etc.); (4) `all'' for comparing all the subsets simultaneously (i.e. \texttt{disparity $\mathtt{\sim}$ subsets}) or (5) any list of pairs (or more) of subsets to compare.

@@

\section{Examples}
Multivariate analysis can be really useful for looking at multiple aspects of organisms diversity together.
For example, one can also look the diversity of morphologies \citep[or disparity;][]{foote1991morphological}.
Using disparity, it is then also possible to assess whether one ecosystem or/and time period display more morphological variation.
The following example is based on a classical morphological disparity analysis.
\hl{Note that more examples are available in the package manual (}\url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}).

@@

In this example, the space is defined as a morphospace: the ordination of the distances among discrete morphological characters for 50 mammal species \citep[from][]{beckancient2014}.
Additionally, we can define disparity as the sum of the variances on each dimension \citep{foote1991morphological,Wills1994} that will represent an aspect of the the volume of the morphospace.
